---
title: "Things that make your life easier in R"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    highlight: kate
    code_download: true
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE,
                      message= FALSE,
                      warning= FALSE)
```

I’ll show some useful packages that may be useful, specially when starting. But before something more advanced available in Rstudio.

## Code snippets

Go to Tools + Global Options + Code + Snippets. We will create a function that extracts to a temporary excel file a data frame.

```{r,eval=FALSE}
show_in_excel
    show_in_excel <- function(.data){
    tmp <- paste0(tempfile(), ".xlsx")
    openxlsx::write.xlsx(.data,tmp)
    browseURL(tmp)
    }
    
show_in_excel(mtcars)
```

We can write a code snippet to call this function every time we need it.

```{r,eval=FALSE}
snippet lb_show_in_excel
    show_in_excel <- function(.data){
    tmp <- paste0(tempfile(), ".xlsx")
    openxlsx::write.xlsx(.data,tmp)
    browseURL(tmp)
    }
    show_in_excel (${1:n})
```

## knitr::purl

Allows to extract a script `(r)` from an rmarkdown `(rmd)` file

```{r,eval=FALSE}
knitr::purl("20211212.rmd")
```

## Datapasta

To easily bring data from internet, excel files, etc. Before copying go to Addins as choose datapasta + paste as tribble

```{r,eval=FALSE}
library(datapasta)

copy<- tibble::tribble(
  ~mpg, ~cyl, ~disp,  ~hp, ~drat,   ~wt, ~qsec, ~vs, ~am, ~gear, ~carb,
    21,   6L,   160, 110L,   3.9,  2.62, 16.46,  0L,  1L,    4L,    4L,
    21,   6L,   160, 110L,   3.9, 2.875, 17.02,  0L,  1L,    4L,    4L,
  22.8,   4L,   108,  93L,  3.85,  2.32, 18.61,  1L,  1L,    4L,    1L,
  21.4,   6L,   258, 110L,  3.08, 3.215, 19.44,  1L,  0L,    3L,    1L,
  18.7,   8L,   360, 175L,  3.15,  3.44, 17.02,  0L,  0L,    3L,    2L,
  18.1,   6L,   225, 105L,  2.76,  3.46, 20.22,  1L,  0L,    3L,    1L
  )
```

## Eurostat

```{r}
library(eurostat)
library(tidyverse)


dd <- get_eurostat("namq_10_gdp", 
                     filters = list(geo = c("EA19","EU27_2020"), 
                                    na_item = "B1GQ", 
                                    unit = "CLV_PCH_PRE",
                                    s_adj = "SCA")) %>% 
  filter(time >= "2008-01-01")
```

Particularly useful is to download the labels of the codes.

```{r, eval = TRUE}
get_label_codes <- function() {
  labels <- tibble(
    dictionary = c("na_item", "nace_r2", "geo", "asset10", "coicop"),
    data = list(
      eurostat::get_eurostat_dic("na_item"),
      eurostat::get_eurostat_dic("nace_r2"),
      eurostat::get_eurostat_dic("geo"),
      eurostat::get_eurostat_dic("asset10"),
      eurostat::get_eurostat_dic("coicop")
    )
  ) %>%
    unnest(cols = c(data))

  return(labels)
}

labels <- get_label_codes()

head(labels)
```
## Esquise

To create a basic ggplot with a GUI. 

```{r, eval=FALSE}
library(esquisse)

esquisse::esquisser(dd)
```

```{r,eval=FALSE}
ddd<-dd %>%
 filter(time >= "2018-10-15" & time <= "2021-07-01") %>%
 ggplot() +
  aes(x = time, fill = geo, weight = values) +
  geom_bar(position = "dodge") +
  scale_fill_viridis_d(option = "plasma", direction = 1) +
  theme(legend.position = "bottom")
```

## ggThemeAssist

To improve ggplot theme elements with a GUI.

```{r,eval=FALSE}
library(ggThemeAssist)

ggThemeAssist::ggThemeAssistGadget(ddd)
```

## Styler

To give a more professional and consistent look to the code style. It is added as an Rstudio Addin.

```{r,eval=FALSE}
library(styler)
```

## Boomer

To view what a long chain of functions do to the data.

```{r}
library(boomer)

dd<- dd %>% 
  pivot_wider(names_from = geo,
        values_from = values) %>% 
  mutate(diff= EA19-EU27_2020) %>% 
  pivot_longer(cols = c(where(is.numeric)),
        names_to = "geo",
        values_to = "values")
```

```{r}
boom(dd<- dd %>% 
  pivot_wider(names_from = geo,
        values_from = values) %>% 
  mutate(diff= EA19-EU27_2020) %>% 
  pivot_longer(cols = c(where(is.numeric)),
        names_to = "geo",
        values_to = "values"))
```

## Tidylog

Similar but less verbose.

```{r message=TRUE}
library(tidylog, warn.conflicts = FALSE)

  dd <- dd%>% 
  filter(geo !="diff") %>% 
  filter(time >="2020-10-01") %>% 
  group_by(geo) %>% 
   pivot_wider(names_from = geo,
        values_from = values)
```

## Janitor

```{r}
library(janitor)
```

### Remove columns with a constant value

```{r}
dd
```

```{r}
dd<- janitor::remove_constant(dd)
dd
```

### Remove duplicates

```{r}
temp<- slice(dd,3)

dd <- bind_rows(dd,temp)

janitor::get_dupes(dd)
```

### Fix column names

```{r}
dd <- janitor::clean_names(dd)
dd
```

### Round like in Excel

```{r}
dd<- dd %>% 
  select(time,ea19) %>% 
  mutate(ea19=ea19 +0.05)

dd
```

```{r}
dd<- dd %>% 
  mutate(normal_round=round(ea19,1),
         excel_round=janitor::round_half_up(ea19,1))
dd
```

