---
title: "Simple outlier detection"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```


## Introduction

Outlier detection can be done using very sophisticated methods. Here I show some very basic approach for practical detection of outliers/errors.


## Data and libraries

```{r}
library(tidyverse)
library(dataregacc)
library(plotly)
```

As data I will use the hours worked per person. That way we can be more efficient and looked at two variables (hours and persons in one shot).

We remove the extra regio.

```{r}
df <- regacc %>% 
  filter(na_item=="EMP" &
         NUTS =="2" &
         country!="MK" &
           !str_ends(geo,"ZZ")&
           values!=0 &
           !nace_r2 %in% c("G-J","K-N","O-U","TOTAL") &
           time>=2000) %>% 
  select(country,geo,nace_r2,unit,time,values) %>%
    pivot_wider(names_from = unit,
              values_from=values) %>% 
  mutate(HW_PS=round(HW/PS)) %>% 
  na.omit()

head(df)
```


## Visualisation


### Distribution

```{r}
ggplotly(ggplot(df,aes(HW_PS, colour=nace_r2))+
  geom_freqpoly(size=1)+
  theme_light())
```

We can already see the distribution of hours worked by nace and noticed that the dispersion for A,L and K is somehow different.

Maybe it is due to small values. If we remove rows that involved less than 5 thousand persons.

```{r}
 ggplotly(ggplot(df %>% filter(PS>=5),aes(HW_PS, colour=nace_r2))+
  geom_freqpoly(size=1)+
  theme_light())
```

### Detailed

But maybe the outliers are due to particular regions. Let's do a more detailed chart.

```{r}
ggplotly(ggplot(df %>% filter(PS>=5),aes(geo,HW_PS))+
  geom_point(colour="darkred")+
  facet_wrap(~nace_r2)+
  theme_light()+
    theme(axis.text.x = element_blank()))
```

This has helped us to identify a few "outliers".

### Data transformation
 
We could transform the data to make other kinds of visualisation.

For time series we could calculate a time mean.

```{r}
df2<-df %>% 
  group_by(country,geo,nace_r2) %>% 
  mutate(time_mean=round(HW_PS*100/mean(HW_PS,na.rm=TRUE)))

head(df2)
```

#### A line plot

```{r}
ggplotly(ggplot(df2 ,aes(time,time_mean, group=geo))+
  geom_line(colour="darkred")+
  facet_wrap(~nace_r2)+
  theme_light())
```

#### A heatmap

Here we will need to restrict a bit, although it is more useful to look at completeness. Let's see only nace L
```{r}
ggplotly(ggplot(df2 %>% filter(nace_r2=="L" & time<=2019) ,aes(time,geo, fill=time_mean))+
  geom_tile()+
  theme_light()+
  scale_fill_gradient(low="darkblue", high = "darkred")+
  scale_x_continuous(expand=c(0,0))+
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank()))
```

