---
title: "Inspect excel files with tidyxl"
author: "Luis Biedma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message = FALSE)
```

## Introduction

Sometimes we are asked to look and review some calculations done in excel or we want to investigate possible (human) errors. The library `{tidyxl}` is very helpful for that task.

## Data and libraries

We are going to look at the central file of the employment estimates. We can load all the cells with the function `xlsx_cells()`. First we create a vector with the names of the sheets, we read them and filter the ones that are numeric. 

```{r}
library(tidyverse)
library(tidyxl)
library(here)

countries <- c(
  "AT", "BE", "BG", "CY", "CZ", "DE", "DK",
  "EE", "ES", "FR", "FI", "GR", "HR", "HU",
  "IE", "IT", "LH", "LU", "LV", "MT", "NL",
  "PL", "PT", "RO", "SE", "SI", "SK"
)
```

## Reading the data

```{r}
df<- tidyxl::xlsx_cells(here("data","central_pers.xlsm")) %>%
  filter(sheet %in% countries &
           !is.na(numeric))

glimpse(df)
```
Let's look at a row to see what information we have.
```{r}
(temp <- df %>% 
  filter(sheet=="BE" & !is.na(formula_ref)) %>% 
  janitor::remove_constant() %>% 
  slice_head(n=10))
```

## Checking formulas

We could, for example, check if for DW6 we are using the same formula in all countries.

```{r}
(temp<- df %>% 
   filter(address=="DW6") %>% 
   select(formula) %>% 
   distinct())
```

But that formula is not very interesting as it only calculates a growth rate. We need to have some knowledge about the file to investigate. We know that rows 117-127 and columns 6-32 contains the annual data of the countries used to calculate the EU aggregates.

```{r}
temp<- df %>% 
   select(sheet, row, col, formula) %>% 
  filter(row >= 117 & row <=127) %>% 
  filter(col>=6 & col <= 32)

head(temp)
```

Now that we have filtered the data  we can check if all the data is sourced the same way. We use `across()` together with `n_distinct()`.

```{r}
distinct<- temp %>% 
  group_by(row,col) %>% 
  summarise(across(c(formula),n_distinct)) %>%
  filter(formula!=1) %>% 
  arrange(desc(formula))

head(distinct)
```

It seems that different formulas are used. Let's look in more detail at row 117 and col 24.

```{r}
temp %>% 
  filter(row=="117" & col=="24" & 
         formula!="VLOOKUP($A117,an!$A$7:$X$1338,COUNTA($B$115:X$115)+1,FALSE)") %>% 
  arrange(formula)
```
SK is linking to a different file and for FR and CY the range is different.
