---
title: "Inset plots with ggplot"
author: "Luis Biedma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message = FALSE)
```

An usual problem we find when making regional maps is what to do with the ultra-peripherical regions. I followed the approach suggested by @dhernangomez (<https://dieghernan.github.io/202203_insetmaps/>).

We will use the following libraries. `{gisco}` for getting the NUTS shapes and `{cowplot}`for arranging the different plots.

```{r}
library(giscoR)
library(sf)
library(tidyverse)
library(cowplot)
```

We get NUTS 2 shapes with a particular epsg which is commonly used for Europe.

```{r}
gisco_set_cache_dir("E:/R/packages/giscoR", install = TRUE, overwrite = TRUE)

regions <- gisco_get_nuts(year= "2021",
                          epsg = "3035", #others ( 4326,3857-Mercator)
                     nuts_level ="2")
ggplot(regions)+
  geom_sf()

```

We set the limits for continental Europe.

```{r}
ggplot(regions)+
  geom_sf()+
  coord_sf(
    xlim = c(2377294, 6500000),
    ylim = c(1413597, 5228510)
  )
```

We create new individual objects transforming the geometry with particular EPSG.
```{r}
canary <- regions %>% 
  filter(NUTS_ID=="ES70") %>% 
  st_transform(4083)

guadeloupe <- regions %>% 
  filter(NUTS_ID=="FRY1") %>% 
  st_transform(4622)

martinique <- regions %>% 
  filter(NUTS_ID=="FRY2") %>% 
  st_transform(4625)

guyanne <- regions %>% 
  filter(NUTS_ID=="FRY3") %>% 
  st_transform(4626)

reunion <- regions %>% 
  filter(NUTS_ID=="FRY4") %>% 
  st_transform(4626)

mayotte <- regions %>% 
  filter(NUTS_ID=="FRY5") %>% 
  st_transform(5793)

malta <- regions %>% 
  filter(NUTS_ID=="MT00")

azores <- regions %>% 
  filter(NUTS_ID=="PT20") %>% 
  st_transform(2188)

madeira <- regions %>% 
  filter(NUTS_ID=="PT30") %>% 
  st_transform(4615)


main <- regions %>% 
  filter(!NUTS_ID %in% c("FRY1", "FRY2", "FRY3", "FRY4", "FRY5", "PT30", "ES70", "PT20", "PT30", "MT00"))
 
```


We create the main plot.

```{r}
main_gg <- ggplot(main) +
  geom_sf() +
# Center in Europe: EPSG 3035
  coord_sf(
    xlim = c(2377294, 6500000),
    ylim = c(1413597, 5228510)
  )+
  theme_void() +
  theme(
    plot.background = element_rect(fill = "grey85", colour = NA),
    # Add a bit of margin on the upper right
    # We would place the inset there
    plot.margin = margin(t = 40, r = 40)
  )
main_gg

```

And the insets for each island.
```{r}
canary <- ggplot(canary) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Canarias (ES)")

canary

malta <- ggplot(malta) +
   geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Malta")

malta

guadeloupe <- ggplot(guadeloupe) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Guadeloupe (FR)")

guadeloupe


martinique <- ggplot(martinique) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Martinique (FR)")

martinique

guyanne <- ggplot(guyanne) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Guyane (FR)")

guyanne

reunion <- ggplot(reunion) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "La Réunion (FR)")

reunion

mayotte <- ggplot(mayotte) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Mayotte (FR)")

mayotte

azores <- ggplot(azores) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Açores (PT)")

azores

madeira <- ggplot(madeira) +
  geom_sf() +
  theme_void() +
  # Add a border to the inset
  theme(
    panel.border = element_rect(fill = NA, colour = "black"),
    plot.background = element_rect(fill = "grey95"),
    plot.title = element_text(size = 6)
  )+
  labs(title = "Madeira (PT")

madeira

```


And we put together all of them with `{cowplot}`. There is a need to play with the different parameters to place the islands where we want.

```{r fig.height=9, fig.width=12}
ggdraw() +
  draw_plot(main_gg) +
  draw_plot(canary,
            height = .1,
            width = .1,
            x = 0.32,
            y = 0.) +
    draw_plot(malta,
            height = .1,
            width = .1,
            x = 0.82,
            y = 0.) +
  draw_plot(guadeloupe,
            height = .1,
            width = .1,       
            x = 0.65,
            y = 0.85,
            scale=2)+
  draw_plot(martinique,
            height = .1,
            width = .1,       
            x = 0.77,
            y = 0.85,
            scale=2)+
  draw_plot(guyanne,
            height = .1,
            width = .1,       
            x = 0.65,
            y = 0.65,
            scale=1.7)+
  draw_plot(reunion,
            height = .1,
            width = .1,       
            x = 0.77,
            y = 0.65,
            scale=1.6)+
  draw_plot(mayotte,
            height = .1,
            width = .1,       
            x = 0.75,
            y = 0.35,
            scale=1.5)+
  draw_plot(azores,
            height = .1,
            width = .1,
            x = 0.12,
            y = 0.65,
            scale=1)+
  draw_plot(madeira,
            height = .1,
            width = .1,
            x = 0.12,
            y = 0.4,
            scale=3)

```





