---
title: "Crosstalk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    highlight: kate
    code_download: true
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE,
                      message= FALSE,
                      warning= FALSE)
```

## Introduction

`{crosstalk}` allow that tables and charts share a dataframe and connect them. It works with `{plotly}`, `{DT}`, `{leaflet}`, `{scatterD3}` and a few other less known libraries.

## Data and libraries

```{r}
library(tidyverse)
library(crosstalk)
library(DT)
library(dataregacc)
library(plotly)
```

```{r}
df<- regacc %>%
  filter(table %in% c("2hh","2gdp","3pop","3emp") &
         na_item %in% c("B1GQ","B6N","EMP","POP") &
          unit %in% c("EUR_HAB", "PS") &
           NUTS=="2" &
           nace_r2 %in% c("TOTAL","Z") &
           time=="2019") %>% 
  select(country,geo,na_item,values) %>% 
  unique() %>% 
    pivot_wider(names_from=na_item,
              values_from=values) %>% 
  na.omit()

```

## Creating a shared object

We create a shared dataset, which it is not a dataframe but a function.

```{r}
shared_df <- SharedData$new(df,key= ~geo)
```

## Linking two charts

We can create two interconnected scatter charts. bscols just puts the two elements horizontally.

```{r}
bscols(
  plot_ly(shared_df, x= ~B1GQ, y =~B6N, text = ~geo),
  plot_ly(shared_df, x=~EMP, y=~POP, text = ~geo)
)
```

We can add filters and slide bars

```{r}
df1<- df %>% 
  mutate(country=str_sub(geo,1,2))

shared_df1 <- SharedData$new(df1,key= ~geo)

bscols(widths = c(3,NA,NA),
  list(
    filter_checkbox("country", "Country", shared_df1, ~country, inline = TRUE),
    filter_slider("B1GQ", "GDP", shared_df1, ~B1GQ, width = "100%")),
  plot_ly(shared_df1, x= ~B1GQ, y =~B6N, text = ~geo,marker=list(size=10,color="red")),
  plot_ly(shared_df1, x=~EMP, y=~POP, text = ~geo, marker=list(size=10,color="blue"))
)
```

## Linking a table and a chart

Or a connected table and chart.

```{r}

  plot_ly(shared_df1, x= ~B1GQ, y =~B6N, text = ~geo,marker=list(size=10,color="red"))
  datatable(shared_df1, filter = "top") 

```

## Linking two tables

or two tables

```{r}

gdp_hh <- df %>%
  select(-POP,-EMP) %>% 
  mutate(country=str_sub(geo,1,2))

emp_pop <- df %>% 
  select(-B1GQ, -B6N) %>% 
  mutate(country=str_sub(geo,1,2))

shared_gdp <- SharedData$new(gdp_hh,key= ~country,group="shared_obj")
shared_emp <- SharedData$new(emp_pop,key= ~country,group="shared_obj")

  filter_select(
  id = "id-selector", label = "Country:", 
  sharedData = shared_gdp, group = ~country)
  
  datatable(shared_gdp, filter = "top") 
  datatable(shared_emp, filter = "top") 

```

