---
title: "List columns and regressions"
author: "Luis Biedma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message = FALSE)
```

## Introduction

We will learn how to use list-columns created with `nest()` (or `split()`), together with `map()` to run many regressions and then gather the results with `broom()`.

## Libraries

We will use the following libraries.

```{r}
library(tidyverse)
library(broom)
library(dataregacc)
```


## Data

We create and tidy the data. We are going to use employees in persons and hours worked by Nace. We will have them in wide format.

```{r}
data <- regacc %>% 
  filter(NUTS==2 &
         na_item=="SAL" &
         table %in% c("3emp","2emphw") &
         !country %in% c("AL","MK", "ME", "RS", "TR", "NO") &
         !str_ends(geo,"ZZ")) %>%
  select(geo,unit,nace_r2,time, values) %>%
  na.omit() %>% 
  pivot_wider(names_from = unit,
        values_from = values)

head(data)
```

## group_by() and nest()
We do a `group_by()` followed by `nest()`.

```{r}
data_nested <- data %>% 
  group_by(geo,nace_r2) %>%
  nest()
```

We have one row by geo and nace and a tibble attached to each row.

```{r}
head(data_nested)
```

Each tibble contains three columns: time, HW and PS

```{r}
head(data_nested$data,1)

```

## Doing the regressions

We can now run the `r nrow(data_nested)` regressions.

```{r}
data_nested <- data_nested %>% 
  mutate(model = map(data, ~ lm(HW ~ PS, data = .x))) 

head(data_nested$model,1)
```

We have now a new column containing the regression model. We now rely in `broom()` to extract more easily the results.

## Broom::tidy()

With `tidy()` we collect information on the parameters, the standard error or the p-value.

```{r}
data_nested<- data_nested %>% 
  mutate(tidy = map(model, tidy))

head(data_nested$tidy,1)

```

## Broom::augment()

With `augment()` we get the fitted results and other information.

```{r}
data_nested<- data_nested %>% 
  mutate(augment = map(model, augment))

head(data_nested$augment,1)
```

## Broom::glance()

With `glance()` we get some summary information like the R^2^, the AIC or the BIC.

```{r}
data_nested<- data_nested %>% 
  mutate(glance = map(model, glance))

head(data_nested$glance,1)
```

We have now all that information together in the dataframe.

```{r}
glimpse(data_nested)
```

## Unnest

Now we could unnest but we have to be careful because some columns appear in several tibbles. We can, for example, keep just the R^2^ and plot the mean .

```{r}
data_unnest<- data_nested %>% 
  select(-model, -tidy,-augment) %>% 
  unnest(cols = c(data,glance)) %>% 
  select(geo,nace_r2,PS,HW,r.squared)
```

```{r}

mean<- data_unnest %>% 
  group_by(nace_r2) %>% 
  summarise(mean=mean(r.squared))

ggplot(mean,aes(mean,fct_reorder(nace_r2,mean)))+
  geom_col(fill="darkred")+
  theme_light()+
  ylab("")+ xlab("")+
  ggtitle("Average R-square between Persons and Hours Worked by NACE")+
  scale_x_continuous(expand=c(0,0))
```

