---
title: "{ggplot2}"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Luis Biedma
output:
  rmdformats::readthedown:
    highlight: kate
    code_download: true
---
```{r setup, include = FALSE}
## Global options
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.height = 9,
  fig.width = 12
)
```

# Introduction 

{Ggplot2} is the default tool for producing charts in R. There are many good books about it and the basic principles are a standard section of most R books. Here I would follow a similar approach as Cédric Scherer and go through the steps of modifying a ggplot (<https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/>).


# Loading the data

```{r}
library(tidyverse)

df<- data.table::fread("data/gdp2v.csv") %>% 
  filter(unit=="PPS_HAB_EU27_2020") %>% 
filter(time == 2018) %>% 
  mutate(country=str_sub(geo,1,2),
         NUTS=str_length(geo)-2) %>% 
  filter(NUTS=="2") %>% 
  select(-obs_decimals,-obs_status, -unit,-NUTS) %>% 
  filter(!country %in% c("AL", "CY", "EE", "LU","LV","ME","MK","MT","NO","RS"))
```


We load the regional GDP as percentage the EU for NUTS 2 regions for the years 2008 and 2018 and we end up with 481 observations and 4 columns.

```{r}
glimpse(df)
```

## A basic plot

We start by defining the dataset to use and the layout. In our case we will have the **country** in the **x** axis and the values in the **y** axis.

```{r}
(ggplot(df, aes(x= country, y= obs_value)))
```

Now we have to tell ggplot how do we want to plot the data. We use `geom_*`for that. Let's say we want to plot dots.

```{r}
g <- ggplot(df, aes(x= country, y= obs_value))

g + geom_point()
```

We could plot a line.

```{r}
g + geom_line()
```

Or a line and a dot.

```{r}
g + geom_point() + geom_line()
```

Within `geom_*`we can control some aesthetics like colour, shape, size.

```{r}
g + geom_point (color="steelblue", size = 3.5, shape = "diamond")
```

Each `geom_` has its own arguments. For example there is no shape for `geom_line`or linetype for `geom_point`.

```{r}
g + 
  geom_point (color="steelblue", size = 3.5, shape = "diamond")+
  geom_line(color="darkred", size = 0.5, linetype = "dashed")

```

Also the order of the geoms (which one shows first) affects the results.
```{r}
g + 
  geom_line(color="darkred", size = 0.5, linetype = "dashed")+
  geom_point (color="steelblue", size = 3.5, shape = "diamond")

```


We can change the default `theme` in ggplot. If we set the theme using `theme_set()` to {theme_bw} we do not have any longer the grey background and other aesthetics also change.

```{r}
theme_set(theme_bw())

g + geom_point(color = "steelblue", size =2.5)
```

You can override a particular thing of a theme, and there are many particular things (<https://ggplot2.tidyverse.org/reference/theme.html>). For example remove the minor gridlines.

```{r}
g + geom_point(color = "steelblue", size =2.5)+
    theme(panel.grid.minor = element_blank())

```

# Axes

## Change Axis title

Let's add some titles to the axes. We can do it with the function `labs`.

```{r}
g + geom_point(color = "steelblue", size =2.5)+
  labs( x = "Country",  y= "As percentage of the EU in PPS")

```

or with `xlab()` and `ylab()`

```{r}
g + geom_point(color = "steelblue", size =2.5)+
  xlab( "Country") +
  ylab( "As percentage of the EU in PPS")

```

We can add symbols, formulas or use superscripts:

```{r}
g + geom_point(color = "steelblue", size =2.5)+
  labs( x = "Country",  y= expression(paste("As percentage of the EU in PPS" ^" EU27 =100")))

```

## Space between Axis and axis titles

We will use `theme()` to modify particular elements (text, background, boxes...). To modify text elements we use `element_text()`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
  labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.title.x = element_text(vjust = 0, size = 15),
        axis.title.y = element_text(vjust = 3, size = 15))
```

`vjust`refers to the vertical alignment.  

We can do the same with `margin()`
```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
  labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.title.x = element_text(margin = margin(t = 10), size = 15),
        axis.title.y = element_text(margin = margin(t = 10), size = 15))
```

`margin()`takes the arguments `t` (top), `r` (right), `b` (bottom) and `l` (left).

## Change Aesthetics of Axis titles

With `axis.title`we can modify the size, colour and face (bold, italic or bold.italic.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.title = element_text(size = 15, color = "steelblue",
                                  face = "italic"))
```

We can do it individually for each axis (this is available in most options).
```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.title.x = element_text(size = 15, color = "steelblue",
                                  face = "italic"),
        axis.title.y = element_text(size = 12, color = "darkred",
                                  face = "bold"))

```

## Change Aesthetics of Axis text

You can change the aesthetics of the axis text with `axis.text`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.text.x = element_text(size = 10, color = "steelblue",
                                  face = "italic"),
        axis.text.y = element_text(size = 12, color = "darkred",
                                  face = "bold"))
```

## Rotate axis text

Specifying an `angle` allows you to rotate any text elements.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.text.x = element_text(angle = 50, size = 12))
```

In that case w will normally need to adjust the distance to the axis.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
  theme(axis.text.x = element_text(angle = 50, size = 12, hjust = 1.5),
        axis.text.y = element_text(angle = 50, size = 12, vjust = -0.5))
```

`hjust` and `vjust` are a bit confusing!

## Remove axis text and ticks

Not very recommended or usual but can be done. Most things are removed using `element_blank()`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs(x = "Country", y = "As percentage of the EU in PPS") +
   theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank())
```

There is `element_text`, `element_line`, `element_rectangle` and `element_blank`. We have seen so far `element_text`and `element_blank`.

## Remove Axis titles

This is more common as sometimes the title of the chart indicates it or is self-evident, like the x axis in our case.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = NULL) 
  
```

Sometimes you do not remember the option *NU* and force it to print nothing with `""`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "") 
```

## Limit the Axis range

Sometimes you want to get a better view at some range of your data. You can do this without filtering your data:

```{r}

ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "") +
    ylim(c(50, 150))
```

Alternatively you can use `scale_y_continuous(limits = c(bottom,top ))` which does the same or `coord_cartesian(ylim = c(bottom, top))` There is a subtle difference between the two, in the second case we really subset the data while in the first we only limit the axis (look at EL and PL!)

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "") +
 coord_cartesian(ylim = c(50, 150))
```

## Coordinates

Let's change a bit our dataset to illustrate the point.
```{r}
df2 <-data.table::fread("data/gdp2v.csv") %>% 
  filter(unit=="PPS_HAB_EU27_2020") %>% 
filter(time == 2008| time == 2018) %>% 
  pivot_wider(names_from = time,
              values_from = obs_value,
              names_prefix = "y") %>% 
  mutate(country=str_sub(geo,1,2),
         NUTS=str_length(geo)-2) %>% 
  filter(NUTS=="2") %>% 
  select(-obs_decimals,-obs_status, -unit,-NUTS) %>% 
  filter(!country %in% c("AL", "CY", "EE", "LU","LV","ME","MK","MT","NO","RS"))
```

Let force it to start at zero.
```{r}
ggplot(df2, aes(y2008,y2018))+
    geom_point(color = "steelblue", size =2.5) +
  expand_limits(x = 0, y = 0)
```

Really at zero!
```{r}
ggplot(df2, aes(y2008,y2018))+
    geom_point(color = "steelblue", size =2.5) +
    expand_limits(x = 0, y = 0) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off")

```
The argument clip = "off" allows to draw outside the panel area.

## Scaling of the axis

We could force the ratio of the axis. The default, ratio = 1, ensures that one unit on the x-axis is the same length as one unit on the y-axis. It can be illustrated better in this chart.

```{r}
ggplot(df2, aes(x = y2008, y = y2018)) +
  geom_point(color = "steelblue", size =2.5) +
    ylim(c(50, 150))+
   xlim (c(20,200))
```

Ratios higher than one in `coord_fixed()`  make units on the y axis longer than units on the x-axis, and vice-versa.

```{r}
ggplot(df2, aes(x = y2008, y = y2018)) +
  geom_point(color = "steelblue", size =2.5) +
coord_fixed(ratio = 1.5)
```

## Use a Function to Alter Labels

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "") +
  scale_y_continuous(label = function(x) {return(paste(x, "pp"))})
```

# Titles

We can add a simple title via the `ggtitle()` function:

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "")+
  ggtitle ("GDP per capita in EU regions")
```

Or use `labs()` and add much more arguments.
```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions",
       subtitle = "Year 2018",
       caption = "Data: Eurostat, nama_10_2gdp",
       tag = "Fig. 1")

```

## Adjust the title 

We use `theme()`as we did with `axis.title` and `axis.text`. The same principles can be used for other elements `plot.subtitle`, `plot.caption`, `plot.caption`, `legend.title` and `legend.text`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions")+
  theme(plot.title = element_text(face = "bold",
                                  margin = margin(10, 0, 10, 0),
                                  size = 16))
```

## Adjust the position of the title

The horizontal adjustment (left,center,right) is controlled by  by `hjust`. For centering we use 0.5 and to the right 1.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions")+
  theme(plot.title = element_text(face = "bold",
                                  margin = margin(10, 0, 10, 0),
                                  size = 16,
                                  hjust =0.5))
```

We can also adjust it vertically with `vjust`.

To move the title outside the panel we can use `plot.title.position` if we have long titles

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions in 2018 as percentage of the EU GDP in Purchasing Power Standards")+
  theme(plot.title.position = "plot")
```

## Change space in multiline text

You can have multiline titles using ´/n`

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions in 2018\n as percentage of the EU GDP in Purchasing Power Standards")
```

And adjust the space between lines with `lineheight`.

```{r}
ggplot(df, aes(x = country, y = obs_value)) +
  geom_point(color = "steelblue", size =2.5) +
   labs( y = "As percentage of the EU in PPS",
         x = "",
         title = "GDP per capita in EU regions in 2018\n as percentage of the EU GDP in Purchasing Power Standards")+ 
  theme(plot.title = element_text(lineheight = 0.7, size = 16))
  
```

